---
title: "STAT 315 Notes"
author: "Colorado State University"
date: ""
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
description: ""
header-includes:
      - \usepackage{pdfpages}
---

# Preliminaries {-}

## Getting Started With R {-}

### Installing R {-}

R is the base version of a free, open-source statistical computing program. R Studio is a nice graphic user interface (GUI) that works in conjunction with R. We will be using R Studio this semester as it is easier to learn how to use and has many nice shortcuts that aren't available in the standard R download.

**Note:** R and R Studio can not be installed on Chromebooks or other tablets/mini-computers. If you only have a tablet or a mini-computer, you can use \href{http://rstudio.cloud/}{R Studio Cloud}

To begin using R Studio, you will need to install "R" first and then install "R Studio" on your computer.


**Step 1: Download R**

- Visit [https://www.r-project.org/](https://www.r-project.org/)
- Click *CRAN* under *Download*
- Select any of the mirrors
- Click the appropriate link for your type of system (Mac, Windows, Linux)
- Download R on this next page.
  - For Windows, this will say *install R for the first time*
  - For Mac, this will be under *Latest release* and will be something like *R-4.1.0.pkg* -- the numbers may differ depending on the most recent version
- Install R on your computer
       
**Step 2: Download R Studio**

- Visit https://www.rstudio.com/products/rstudio/download/#download
- Click to download
- Install R Studio on your computer

\newpage

**Step 3: Verify R Studio is working**

- Open R Studio
- Let's enter a small dataset and calculate the average to make sure everything is working correctly.
- In the console, type in the following dataset of five homework scores:

```{r}
# Assignment using = 
x = c(90,75,100,85,90)

# Alternative assignment using <-
x <- c(90,75,100,85,90)

# More on the differences between = and <- at:
# https://www.r-bloggers.com/2014/01/difference-between-assignment-operators-in-r/
```

In the console, calculate the average of the five homework scores: \

```{r}
mean(x)
```

Did you find the average of the five homework scores was 88? If so, you should be set up correctly!